Characterization of gaps in directly bonded Si compound optics

### 2014-04-21 AO_bondedSi.tex
minor tweaks to bubble-gap-void word usage

While it is possible that proper precautions to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap, exceeding the scale over which interfacial molecular forces are acting, roughly $>$2 nm.

@@

There are two main solutions for reducing risk attributable to gaps.  The first approach is to prevent bubbles from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach involves disciplined handling protocols, and rigorous multi-stage cleaning steps.  The reason the approach is so hard is that hydrocarbons can serve as catalysts for gap formation\cite{1991JaJAP..30..615M}, and that these hydrocarbons are common in Si wafer handling equipment.  Unconventional micro-cleanroom approaches have been effective at meeting the goal of bubble-free bonding\cite{1989JaJAP..28L2141L}.  The second approach is to accept the likelihood that gaps will occur, and take steps to remove them.  The most cited way to get rid of gaps is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in Section \ref{secHistory}.

@@

Consider the following two conceivable pitfalls in Si-Si direct bonding.  The obvious pitfall is particle contamination.  Figure \ref{figParticle} shows a schematic of a particle wedged between two flat Si optical surfaces.  A single 0.6 $\mu$m particle can create a dark IR bubble from interference.  In the next section we show that smaller bubbles can unacceptably degrade the performance to as low as 35 nm.  In a class 100 cleanroom, the volume density could be as high as $\rho=100$ particles/ft$^3$, with sizes larger than 0.5 $\mu$m.  What we care about is not the volume number density of particles, but the delivered surface flux $F=\rho(D) v(D)$, where $v(D)$ is the average freefall speed of particles, which is a function of particle size.  Figure 5 of Cooper 1986\cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^{-2}$ s$^{-1}$) vs. particle size ($\mu$m).  For a change from 1 $\mu$m to 10$^{-2}$ $\mu$m, there is an increase in a factor of $10^4$ in the particle flux expected.  In other words:
$$\log(F) = -2\log(D) + const.$$

@@

It has been shown\cite{1992JEMat..21..669M} that the gases seep through the Si$-$Si interface.  Pressure builds around catalysts.  In this sense, the size of the particle causing a gap cannot be uniquely identified from observations of the extent of gaps.  In section \ref{secResults} we address the question of the origin of the IR bubbles.  Specifically, are the particles more like \emph{tentpoles}, rigid structures that merely prevent the zippering up of the bond?  Or are the particles more like \emph{catalysts} so that their size is immaterial- the particles merely accelerate the production of gases that bow out the bond?  These are open questions.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with the coarse IR imaging technique.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  Figure \ref{IRbubble} shows an example of one of the IR bubbles in which several bright and dark fringes are clearly discernible.  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\section{Results of infrared spectroscopy of bonded Si pucks}
\label{secResults}
In this section we describe the spectroscopy of bonded Si pucks.  In some instances IR imaging was available and was leveraged to target IR bubbles.  In other cases the IR imaging was not available or simply ignored, and spectra were taken at random positions on the wafer surface.

@@

\subsection{Mitigation of measurement artifacts}
The principle artifacts and their interpretation are listed in Table \ref{tabInterpretation}.  We mitigated the artifacts in the following ways.  For the Sharpie marker behaving as an anti-reflection (AR) coating, we simply removed the sharpie with IPA.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature \cite{1992JEMat..21..669M}.  Specifically, the gas pressure of bonded wafers grows as the fill factor of voids decreased.  In other words, what matters is the fill factor of voids, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the voids, and therefore no difference in the bonding properties.

@@

\section{Conclusions}
We have shown a method for indirectly measuring the size of voids in bonded Si wafers.  The technique broadly applies to any compound Si optics measured at normal incidence.  We showed lab IR spectra with consistent predictions, and addressed ways to mitigate measurement artifacts.

>>>>

While it is possible that proper precautions to eliminate all micron-sized particles from both Si surfaces prior to bonding, the power-law nature of the particle size distribution in clean-room environments \cite{doi:10.1080/02786828608959094} means that it is significantly more difficult to assure that no particles of the 10$-$60 nm size scale are present.  Small particles could be a concern if they act as "tent poles", that create gaps where the bonding is incomplete (Figure \ref{figParticle}).  Some authors \cite{1991JaJAP..30..615M, 1992JEMat..21..669M, 1989JaJAP..28L2141L, Mitani1990} have referred to these large ($\sim$ 1 $\mu$m size) gaps as \emph{IR bubbles}, or simply bubbles, owing to their appearance in infrared (IR) transmission images as contiguous regions of a deficit of IR transmission.  The gaps have also been referred to as voids\cite{2000RScI...71.1869G}, interfacial voids, or gaps.  In this paper we will reserve the term ``IR bubbles'' for gaps that are large enough to be detected directly in conventional IR imaging.  The term ``gap'' will apply more broadly to any finite size gap, exceeding the scale over which interfacial molecular forces are acting, roughly $>$2 nm.  We omit the adjective ``interfacial'' before gap for simplicity.

@@

There are two main solutions for reducing risk attributable to gaps.  The first approach is to prevent gaps from forming in the first place, specifically by removing all possible contaminants from the environment before bonding.  Typically this approach involves disciplined handling protocols, and rigorous multi-stage cleaning steps.  The reason the approach is so hard is that hydrocarbons can serve as catalysts for gap formation\cite{1991JaJAP..30..615M}, and that these hydrocarbons are common in Si wafer handling equipment.  Unconventional micro-cleanroom approaches have been effective at meeting the goal of bubble-free bonding\cite{1989JaJAP..28L2141L}.  The second approach is to accept the likelihood that gaps will occur, and take steps to remove them.  The most cited way to get rid of gaps is to anneal them\cite{Mitani1990}.  We describe the current understanding of both approaches in Section \ref{secHistory}.

@@

Consider the following two conceivable pitfalls in Si-Si direct bonding.  The obvious pitfall is particle contamination.  Figure \ref{figParticle} shows a schematic of a particle wedged between two flat Si optical surfaces.  A single 0.6 $\mu$m particle can create a dark IR bubble from interference.  In the next section we show that small gaps can unacceptably degrade the performance to as low as 35 nm.  In a class 100 cleanroom, the volume density could be as high as $\rho=100$ particles/ft$^3$, with sizes larger than 0.5 $\mu$m.  What we care about is not the volume number density of particles, but the delivered surface flux $F=\rho(D) v(D)$, where $v(D)$ is the average freefall speed of particles, which is a function of particle size.  Figure 5 of Cooper 1986\cite{doi:10.1080/02786828608959094} shows the surface flux (particles cm$^{-2}$ s$^{-1}$) vs. particle size ($\mu$m).  For a change from 1 $\mu$m to 10$^{-2}$ $\mu$m, there is an increase in a factor of $10^4$ in the particle flux expected.  In other words:
$$\log(F) = -2\log(D) + const.$$

@@

It has been shown\cite{1992JEMat..21..669M} that the gases seep through the Si$-$Si interface.  Pressure builds around catalysts.  In this sense, the size of the particle causing a gap cannot be uniquely identified from observations of the extent of gaps.  In section \ref{secResults} we address the question of the origin of interfacial gaps.  Specifically, are the particles more like \emph{tentpoles}, rigid structures that merely prevent the zippering up of the bond?  Or are the particles more like \emph{catalysts} so that their size is immaterial- the particles merely accelerate the production of gases that bow out the bond?  These are open questions.

@@

We looked for the indication of large gaps detectable as ``IR bubbles'' \cite{1992JEMat..21..669M} with IR imaging.  The detector was an IR Vista $\alpha$NIR infrared focal plane array, with 312 $\times$ 252 pixels, with 30 $\mu$m square pixels.  We used a 50 mm F/2 lens, with a fiber optic white light illumination source.  We estimate that the lamp-detector combination was sensitive to wavelengths in the range 1.1-1.6 $\mu$m.  The field of view sub-sampled the 4 inch diameter wafers, so we dithered the sample.  No effort was made to flat field the detector, so copious detector non-uniformities and vignetting are present in our images.  It was easy to detect IR bubbles despite the detector artifacts since the bubbles translate with the wafer during dithering.  We detected IR bubbles in all 3 of the bonded Si pucks on which we recorded IR images.  The bubble areal density varied from about 4 per 100 mm diameter bonded wafer pair to over 20 per bonded wafer pair.  Naturally, the unbonded DSP wafers show no such IR bubbles, which are entirely attributable to the gaps between bonded substrates as discussed in the literature (see Section \ref{secHistory}).  Figure \ref{IRbubble} shows an example of one of the IR bubbles in which several bright and dark fringes are clearly discernible.  No attempt at quantitative measurement was made on the IR images due to evidence that the detector is non-linear.  The morphology was clear regardless of the detector response.

@@

\section{Results of infrared spectroscopy of bonded Si pucks}
\label{secResults}
In this section we describe the spectroscopy of bonded Si pucks.  In some instances IR imaging was available and was leveraged to target or avoid locations demonstrating IR bubbles.  In other cases the IR imaging was not available or ignored, and spectra were taken at random positions on the wafer surface.

@@

\subsection{Mitigation of measurement artifacts}
The principle artifacts and their interpretation are listed in Table \ref{tabInterpretation}.  We mitigated the artifacts in the following ways.  For the Sharpie marker behaving as an anti-reflection (AR) coating, we simply removed the sharpie with IPA.  We have mostly been able to rule out wavelength-dependent diffraction from the mesh pattern as a measurement artifact, since the beam size is large and the angular spread of diffraction from the mesh pattern would be much smaller than the spread of the beam.  Still, to remove this concern, we moved to using the coarse mesh rather than the medium or fine meshes, when possible.  We saw no difference in the bond quality surrounding the medium or fine meshes, which is consistent with results from the literature \cite{1992JEMat..21..669M}.  Specifically, the gas pressure of bonded wafers grows as the fill factor of gaps decreased.  In other words, what matters is the fill factor of gaps, not their relative sizes.  Since the fine, medium, and coarse meshes had the same fill factors, we anticipate no difference in the pressure of the gaps, and therefore no difference in the bonding properties.

@@

\section{Conclusions}
We have shown a method for indirectly measuring the size of gaps in bonded Si wafers.  The technique broadly applies to any compound Si optics measured at normal incidence.  We showed lab IR spectra with consistent predictions, and addressed ways to mitigate measurement artifacts.
